---
title: "Getting Started with ccreds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ccreds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette walks through a complete `ccreds` workflow: simulating
censored-covariate regression data, fitting the model with
cross-validated penalty selection, and visualising the results.

## Setup

```{r setup}
library(ccreds)
library(MASS)
library(dplyr)
```

## Simulate data

We generate data from the censored-covariate regression model with
$n = 200$ observations and $p = 10$ features.

**Response-variable model:**
$$y_i = \beta_0 + \alpha x_i + \pmb{\beta}^\top \mathbf{z}_i + \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)$$

**Censored-covariate model:**
$$X_i \mid \mathbf{z}_i \sim \chi^2(\mu_i), \quad \mu_i = \exp(\gamma_0 + \pmb{\gamma}^\top \mathbf{z}_i)$$

We set $\pmb{\beta}$ and $\pmb{\gamma}$ to be sparse with
partially overlapping support, so that some features affect only the
response, some affect only the censored covariate, and some affect both.

```{r simulate}
set.seed(8371)

n <- 200
p <- 10

# Correlated features (AR(0.5) structure)
rho <- 0.5
Sigma <- 9 * toeplitz(rho^(0:(p - 1)))
z <- mvrnorm(n = n, mu = rep(0, p), Sigma = Sigma)

# True parameters
beta_0 <- 5
alpha <- 10
beta <- c(4, 4, 2, 0, 0, 0, 0, 0, -2, -4)
sigma_sq <- 100

gamma_0 <- 2
gamma <- c(0, 0.04, 0.04, 0.02, 0.02, 0, 0, 0, 0, 0)

# Generate the censored covariate x
x_mean <- exp(cbind(1, z) %*% c(gamma_0, gamma)) |> drop()
x <- rchisq(n, df = x_mean)

# Generate censoring times to achieve ~50% censoring
compute_c_mean <- function(censoring_rate, x_mean) {
  f <- function(c_mean) {
    1 - pbeta(0.5, x_mean / 2, c_mean / 2) - censoring_rate
  }
  uniroot(f, lower = 1e-12, upper = 1000)$root
}
c_mean <- vapply(x_mean, \(m) compute_c_mean(0.5, m), numeric(1))
cens <- rchisq(n, df = c_mean)

# Generate response
e <- rnorm(n, mean = 0, sd = sqrt(sigma_sq))
y <- cbind(1, x, z) %*% c(beta_0, alpha, beta) + e

# Observed covariate and censoring indicator
w <- pmin(x, cens)
d <- as.numeric(x <= cens)

# Assemble the data frame in the format ccreds() expects
z_df <- as.data.frame(z)
colnames(z_df) <- paste0("z", 1:p)
dat <- tibble(
  case = 1:n,
  y = drop(y),
  w = w,
  d = d
) |>
  bind_cols(z_df)

dat
```

```{r cens-rate}
mean(dat$d == 0)
```

## Fit the model

We search over a small grid of penalty parameters. In practice you would
use a finer grid and potentially run in parallel with `cores > 1`.

```{r fit, message = FALSE}
lambda_1s <- c(0.01, 0.05, 0.1, 0.5, 1)
lambda_2s <- c(0.01, 0.05, 0.1, 0.5, 1)

results <- ccreds(
  data = dat,
  lambda_1s = lambda_1s,
  lambda_2s = lambda_2s,
  k = 5
)

results
```

## Identify the best model

Use `extract_fit()` to retrieve the fitted model. The default
`rule = "min"` selects the $(\lambda_1, \lambda_2)$ pair that
minimises `CV_MAE_y`. The `rule = "1se"` alternative selects the most
parsimonious model whose `CV_MAE_y` is within one standard error of
the minimum.

```{r best-model}
fit <- extract_fit(results)
```

We can inspect the estimated coefficients and compare them to the truth:

```{r coefficients}
# Response-model coefficients
tibble(
  feature = paste0("z", 1:p),
  true_beta = beta,
  estimated_beta = round(fit$beta, 3)
)

# Censored-covariate coefficients
tibble(
  feature = paste0("z", 1:p),
  true_gamma = gamma,
  estimated_gamma = round(fit$gamma, 5)
)
```

## Visualise: Cross-validation heatmap

`plot_cv()` shows how the CV metric varies over the penalty grid.
The white cross marks the optimal pair.

```{r plot-cv, fig.alt = "Heatmap of cross-validated MAE over the lambda grid."}
plot_cv(results)
```

You can also plot `CV_MAE_x` (prediction error for the censored
covariate among complete observations):

```{r plot-cv-x, fig.alt = "Heatmap of cross-validated MAE for x."}
plot_cv(results, metric = "CV_MAE_x")
```

## Visualise: Coefficient paths

`plot_coefficient_paths()` shows how the estimated coefficients change
as the penalty parameter varies, holding the other penalty fixed at
the value from the best model.

```{r coef-paths-beta, fig.alt = "Coefficient paths for beta as lambda_1 varies."}
plot_coefficient_paths(results, component = "beta", highlight = TRUE)
```

```{r coef-paths-gamma, fig.alt = "Coefficient paths for gamma as lambda_2 varies."}
plot_coefficient_paths(results, component = "gamma", highlight = TRUE)
```

## Visualise: Predictive densities

`plot_density()` displays the joint or conditional density of
$(y, x)$ for a given feature vector $\mathbf{z}$.

```{r density-joint, fig.alt = "Joint density contour plot of y and x."}
z_new <- rep(0, p)
plot_density(fit, z = z_new)
```

Supplying `y_given` shows the marginal density $f(x \mid \mathbf{z})$
alongside the conditional density $f(x \mid \mathbf{z}, y)$:

```{r density-conditional, fig.alt = "Marginal vs conditional density of x given z and y."}
plot_density(fit, z = z_new, y_given = 50)
```

## Summary

The typical `ccreds` workflow is:

1. **Prepare data** in the required format: `case`, `y`, `w`, `d`,
   `z1`, ..., `zp`.
2. **Fit** with `ccreds()` over a grid of $(\lambda_1, \lambda_2)$
   values.
3. **Select** the best model with `extract_fit()`.
4. **Visualise** with `plot_cv()`, `plot_coefficient_paths()`, and
   `plot_density()`.
